---
title: "DUNL Morphomtric Analysis"
author: "Gianco AngelozziBlanco"
date: "2024-03-29"
output: html_document
---

## Tringa Analysis: these data come from Ruthrauff, D.R., Tibbitts, T.L., Gill, Jr., R.E., Handel, C.M., 2023, Adult shorebird morphological measurement data (ver 3.0, November 2024): U.S. Geological Survey data release, https://doi.org/10.5066/P9KNRWXB
```{r}
library(tidyverse)
library(ggpubr)
library(emmeans)
library(DHARMa)
library(car)
alaska_shorb_data<-read_csv("adultShorebird_morphologicalMeasurements_values.csv")

tringa_data=alaska_shorb_data%>%filter(Common_Name%in%c("Greater Yellowlegs", "Lesser Yellowlegs"),,
         !is.na(Diagonal_Tarsus),
         !is.na(Exposed_Culmen))%>% ##remove birds that had no measurements taken
  mutate(bt_ratio=Exposed_Culmen/Diagonal_Tarsus)%>%
  rename("species"=Common_Name)%>%##rename Common name to species
  filter(!bt_ratio>1)## filter 2 outliers

tringa_data%>%mutate(species=as.factor(species))%>%
  ggplot(aes(x = species, y=bt_ratio))+
  geom_boxplot(aes(x = species, y=bt_ratio,colour=species))+
  geom_jitter(aes(x = species, y=bt_ratio, colour=species))+scale_colour_viridis_d(direction = -1, begin = .4, end = .8)+
  ylab("bill-to-tarsus ratio")+
  ggpubr::theme_pubr(base_size = 14) ## 2 values for LEYE are way out of range


lm_ratio<-lm(bt_ratio~species, data = tringa_data)
simulateResiduals(lm_ratio, plot=T) #no issues with variance or dispersion
summary(lm_ratio)
Anova(lm_ratio, type = "II") #type II ss for degrees of freedom (unequal sample sizes suspected). Ratios differ by species

emmeans(lm_ratio, specs = "species") ## mean for GRYE 0.94, mean for LEYE 0.67, no CI overlap at all

contrast(emm, method = "pairwise")%>%as.data.frame() #the mean difference is 0.23 in ratio scale between species

emm <- emmeans(lm_ratio, ~ species) %>%
  as.data.frame()

ggplot(emm, aes(x = species, y = emmean)) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +  # CI
  geom_point(size = 4, aes(colour=species)) +  # Mean estimate
  labs(x = "Group", y = "Estimated Mean", title = "Estimated Marginal Means with 95% CIs") +scale_colour_viridis_d(direction = -1, begin = .4, end = .8)+
  ggpubr::theme_pubr(base_size = 14) ##final plot for the differences between species.

```
## Calidris alpina analysis

## Generate distributions for tarsus and bill - arctica (Greenwood 1979)

```{r}
library("DescTools")
set.seed(100)

arct_male<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 22.8, sd=.64), ##simulation male arctica
         rdist2 = rnorm(n = 50, mean = 26.1, sd=1.1))
subspp<-paste0(rep("arctica", each=50))
sex<-paste0(rep("male"))
x1<-cbind(subspp, sex)
  
arct_female<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 23.4, sd=.91), ##simulation female arctica
         rdist2 = rnorm(n = 50, mean = 29.5, sd=1.09))
subspp<-paste0(rep("arctica", each=50))
sex<-paste0(rep("female"))
x2<-cbind(subspp, sex)

act_sex<-rbind(x1,x2)


arctica<-rbind.data.frame(arct_male, arct_female)

Dunl1<-cbind(arctica, act_sex)

```
##Generate distributions for tarsus and bill - schinzii (Greenwood 1979)

```{r}
schinz_male<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 24.1, sd=.84), ##male schinzii
         rdist2 = rnorm(n = 50, mean = 28.7, sd=1.74))
subspp<-paste0(rep("schinzii", each=50))
sex<-paste0(rep("male"))
x1<-cbind(subspp, sex)
  
schinz_female<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 25.1, sd=.92), ##female schinzii
         rdist2 = rnorm(n = 50, mean=31.7, sd=2.05))
subspp<-paste0(rep("schinzii", each=50))
sex<-paste0(rep("female"))
x1<-cbind(subspp, sex)

sch_sex<-rbind(x1,x2)
sch<-rbind.data.frame(schinz_male, schinz_female)

Dunl2<-cbind(sch, sch_sex)

#tars_shin_male<-rnorm(50, mean = 24.1, sd=.84)
#bill_shin_male<-rnorm(50, mean = 28.7, sd=1.74)
#tars_shin_female <-rnorm(50, mean = 25.1, sd=.92)
#bill_shin_female<-rnorm(50, mean = 31.7, sd=2.05)
```
## Generate distributions for tarsus and bill - hudsonia

```{r}
huds_male<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 27.39, sd=.06),
         rdist2 = rnorm(n = 50, mean = 48.96, sd=2.43))
subspp<-paste0(rep("husdonia", each=50))
sex<-paste0(rep("male"))
x1<-cbind(subspp, sex)
  
huds_female<-RndPairs(50, r=0, rdist1 = rnorm(n = 50, mean = 27.44, sd=.76),
         rdist2 = rnorm(n = 50, mean = 50.26, sd=2.44))
subspp<-paste0(rep("husdonia", each=50))
sex<-paste0(rep("female"))
x2<-cbind(subspp, sex)

hud_sex<-rbind(x1,x2)
hud<-rbind.data.frame(huds_male, huds_female)

Dunl3<-cbind(hud, hud_sex)


DUNL_test<-rbind.data.frame(Dunl1, Dunl2, Dunl3)
names(DUNL_test) <- c("tarsus", "bill", "subspp", "sex")
write.csv(DUNL_test, "DUNLsim2.csv", row.names = F) ##This is the exact file I worked out below.


```

##Read dataset
```{r}
library(tidyverse)

dunl<-read.csv("DUNLsim.csv")

dunl<-dunl[,1:4] #get rid of garbage variables and rows
dunl$btot<-dunl$bill/dunl$tarsus
dunl$sex<-as.factor(dunl$sex)
dunl$subspp<-as.factor(dunl$subspp)

str(dunl)
dunl$btot<-round(dunl$btot, digits = 2)

dunl<-dunl%>%
  filter(!is.na(bill))

dunl%>%
  group_by(subspp, sex)%>%
  summarise(median=median(btot), mean=mean(btot), sd=sd(btot), min=min(btot), max=max(btot))%>%
  ggplot( aes(x=subspp, y=mean, colour=sex)) +  
  geom_point(aes(colour=sex))+ 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2, 
                position=position_dodge(0.0))+
  geom_hline(yintercept = 1.16)+
  xlab("Subspecies")+
  ylab("bill-to-tarsus ratio")


summary_stats_bySpSex<-dunl%>%
  group_by(subspp, sex)%>%
  summarise(mean=mean(btot), sd=sd(btot), min=min(btot), max=max(btot))

```
## Getting Z scores
```{r} 
calculate_z <- function(X, X_mean, S){
  return((X-X_mean)/S)
}

dunl$z_score <- calculate_z(dunl$btot, 
                             mean(dunl$btot, na.rm = TRUE),
                             sd(dunl$btot, na.rm = TRUE))

theme_coding_the_past <- function() {
  theme_bw()+
  theme(# Changes panel, plot and legend background to dark gray:
        panel.background = element_rect(fill = '#2E3031'),
        plot.background = element_rect(fill = '#2E3031'),
        legend.background = element_rect(fill="#2E3031"),
        # Changes legend texts color to white:
        legend.text =  element_text(color = "white"),
        legend.title = element_text(color = "white"),
        # Changes color of plot border to white:
        panel.border = element_rect(color = "white"),
        # Eliminates grids:
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        # Changes color of axis texts to white
        axis.text.x = element_text(color = "white"),
        axis.text.y = element_text(color = "white"),
        axis.title.x = element_text(color="white"),
        axis.title.y = element_text(color="white"),
        # Changes axis ticks color to white
        axis.ticks.y = element_line(color = "white"),
        axis.ticks.x = element_line(color = "white")
  )
}

summ_stats_all<-dunl%>%
  select(subspp, btot)%>%
  group_by(subspp)%>%
  summarise(mean=mean(btot), sd=sd(btot))
  

summ_stats_hudsonia<-dunl%>%
  select(subspp, sex, btot)%>%
  group_by_("sex")%>%
  filter(subspp=="hudsonia")%>%
  summarise(mean=mean(btot), sd=sd(btot))

summ_stats_hudsonia

meanF<-summ_stats_hudsonia$mean[1]
sdF<-summ_stats_hudsonia$sd[1]

meanM<-summ_stats_hudsonia$mean[2]
sdM<-summ_stats_hudsonia$sd[2]

hud_f=dunl%>%
  select(subspp,sex, btot)%>%
  filter(subspp=="hudsonia" & sex=="female")%>%
ggplot(aes(x = btot))+
  geom_histogram(fill = "grey55", 
                color = "#FF6885", 
                alpha = 0.6, 
                bins = 20)+
  ylab("Count")+
  xlab("Bill-to-Tarsus Ratio")+
  geom_vline(xintercept = meanF + 2*sdF, 
             color = "black",   
             linetype = "dashed")+
  geom_vline(xintercept = meanF - 2*sdF, 
             color = "black",   
             linetype = "dashed")+
  geom_vline(xintercept = meanF + 3*sdF, 
             color = "red", 
             linetype = "dashed")+
  geom_vline(xintercept = meanF - 3*sdF, 
             color = "red", 
             linetype = "dashed")+
  annotate("text", x = 1.62,  y = 20, 
            label = "Z=-3", color = "#feb24c")+
  annotate("text", x = 2.2,   y = 20, 
            label = "Z=3",  color = "#feb24c")+
  annotate("text", x = 1.68,  y = 20, 
            label = "Z=-2", color = "red")+
  annotate("text", x = 1.93, y = 20, 
            label = "Z=2",  color = "red")+
  theme_minimal()+ggtitle("Z-scores for female hudsonia") ## histogram for Z-scores of female hudsonia

hud_m=dunl%>%
  select(subspp,sex, btot)%>%
  filter(subspp=="hudsonia" & sex=="male")%>%
ggplot(aes(x = btot))+
  geom_histogram(fill = "grey55", 
                color = "#FF6885", 
                alpha = 0.6, 
                bins = 20)+
  ylab("Count")+
  xlab("Bill-to-Tarsus Ratio")+
  geom_vline(xintercept = meanM + 2*sdM, 
             color = "black",   
             linetype = "dashed")+
  geom_vline(xintercept = meanM - 2*sdM, 
             color = "black",   
             linetype = "dashed")+
  geom_vline(xintercept = meanM + 3*sdM, 
             color = "red", 
             linetype = "dashed")+
  geom_vline(xintercept = meanM - 3*sdM, 
             color = "red", 
             linetype = "dashed")+
  annotate("text", x = 1.62,  y = 20, 
            label = "Z=-3", color = "red")+
  annotate("text", x = 1.9,   y = 20, 
            label = "Z=3",  color = "red")+
  annotate("text", x = 1.68,  y = 20, 
            label = "Z=-2", color = "black")+
  annotate("text", x = 1.93, y = 20, 
            label = "Z=2",  color = "black")+
  theme_minimal()+ggtitle("Z-scores for male hudsonia") ## z-scores for male hudsonia

print(summary_stats_bySpSex)

Dunl_coche<-2.20/2 ##Bill to Tarsus ratio from Coche Dunlin (in pixel units)
Dunl_coche

##Measuring variability in BtoT ratio- these come from the photograps taken by Sabino 

a<-(3.5/3.2) ##trial photo 1
b<-(5.8/5.1)##trial photo 2
c<-(4.4/3.9)##trial photo 3
rbind(a,b,c)
3.359/3

sd(rbind(a,b,c)) ##the most amazing SD, very small.

  ##Z-scores per subspp and sex- score from coche-mean score/sd for the sex x race
far<-(Dunl_coche-1.2398)/0.07092652 #ArctFem
mar<-(Dunl_coche-1.1288)/0.05457218 #ArctMal
fsc<-(Dunl_coche-1.2572)/0.11243211 #SchiFem
msc<-(Dunl_coche-1.1910)/0.07754525 #SchiMal
fhu<-(Dunl_coche-1.8262)/0.04494396 #HudFem
mhu<-(Dunl_coche-1.7832)/0.07851790 #HudMal

far #Z-scores by subspp x sex combo
mar
fsc
msc
fhu
mhu

#p-values for the comparisons
round((1-pnorm(abs(far)))*2, digits = 2)
(1-pnorm(abs(mar)))*2
(1-pnorm(abs(fsc)))*2
(1-pnorm(abs(msc)))*2
(1-pnorm(abs(fhu)))*2
(1-pnorm(abs(mhu)))*2

```
##Graph Subspp by Sex + Sabino's DUNL
```{r}
library(ggpubr)
library(tidyverse)


DNLN_box<-dunl%>%
  ggplot(aes(x=fct_reorder(subspp, btot, .fun = median,.desc = F), y=btot, fill=sex))+
  geom_boxplot(alpha=.5)+
scale_fill_grey()+
  geom_hline(yintercept = 1.1, show.legend = T, linetype=2, colour="blue")+
  xlab("Dunlin Subspecies")+
  ylab("Bill to Tarsus Ratio")+
  theme(legend.position = "right")+
  theme_pubr(border = T, legend = "right")+
  annotate("text", x=1.42, y=1.43, label="Most likely subspecies", size=3)+ 
  annotate("rect", xmin = 0.6, xmax = 2.4, ymin = 1.02, ymax = 1.45,
  alpha = .15)+ theme(axis.text.x = element_text(face = "italic"))+ 
  theme(legend.position = c(0.092, 0.80))+
  theme(legend.background = element_rect(fill="gray95", 
                                  size=0.5, linetype="solid"))
DNLN_box

ggsave("DUNL.jpeg", plot = DNLN_box, dpi = 350, device = "jpeg", units="cm",width = 15, height = 7.5)
```
